---
title: "Final Assignment"
author: "Ken Markus"
date: "November 27, 2015"
output: html_document
---

This project is where you show off your ability to (1) use R packages, (2) change the shape of data in a data frame, and (3) provide basic summary statistics and graphics as part of your exploratory data analysis.
* You should start by installing and loading the R package hflights, which is available on CRAN and described
here: https://cran.r-project.org/web/packages/hflights/hflights.pdf
* You should perform some basic transformations on the data frame, such as including only a subset of columns,
created one or more new derived columns (such as flight date), and/or filtering out rows.
* You should then create basic summary statistics. You should create at least one interesting question for
analysis, and answer the question using two or more graphics in R.

```{r include=FALSE}
# Load the flights package along with a few others we know we'll need
require(hflights)
require(ggplot2)
require(stringr)
require(plyr)
```

```{r}
# Let's take a look at the data
head(hflights)

# Now let's look at the summary information
summary(hflights)
```

Lots of interesting information here. For purposes of the final assignment, we will focus on analytics around *delays*. Ideally identifying some predictability around delays so we can be smarter about our bookings.

```{r}
#create a new dataframe removing the non-finite Delay and Arrival values
finite_index = is.finite(hflights$DepDelay) == TRUE
hflights_usable <- hflights[finite_index, ]

finite_index = is.finite(hflights_usable$ArrDelay) == TRUE
hflights_usable <- hflights_usable[finite_index, ]

#Let's look at the chart
ggplot(data = hflights_usable) + geom_density(aes(x = DepDelay))
```

One thing that is quite clear visually, is that the delays exhibit a 'long tail' like distribution. If we take a look at the percentile distributions, we will see further evidence of large values away from the mean.
```{r}
quantile(hflights_usable$DepDelay, c(0.1, .2, .3, .4, .5, .6, .7, .8, .9, .99, .999))
```

We will also edit the data frame such that we can see relationships of some variables (delays) relative to the date of the delay.
```{r}
hflights_dates <- paste(hflights_usable$Year, "-", hflights_usable$Month, "-", hflights_usable$DayofMonth)
# Remove all of the whitespace
hflights_dates <- str_replace_all(hflights_dates, " ", "")

#convert the strings to Date class
hflights_dates <- as.Date(hflights_dates)


#append the date information to our usable data frame, and remove the non-essential date information
hflights_usable[,"FlightDate"] <- hflights_dates
hflights_usable <- subset(hflights_usable, select=-Year)
hflights_usable <- subset(hflights_usable, select=-Month)
hflights_usable <- subset(hflights_usable, select=-DayofMonth)
```

For our first visual data analysis, let's see if delays exhibited any pattern or fluctuation over time
```{r}
ggplot(hflights_usable, aes(x = FlightDate, y = DepDelay)) + geom_point(shape=1)
```

Upon review we see **no pattern** here. So we'll need to dive in more.

Next, let's take a look at the relationship between departure delays and arrival delays, likely there is a relationship...
```{r}
ggplot(hflights_usable, aes(y = DepDelay, x = ArrDelay)) + geom_point(aes(color = Origin))
```

Voila! Clearly there is a relationship, and arrival delays can be predictive when it comes to departure delays.

Perhaps the carrier also has some effect. Now, we'll focus on arrival delays.
```{r}
ggplot(hflights_usable, aes(x = UniqueCarrier, y = ArrDelay)) + geom_violin()
```

This is indeed the case, with certain carriers, such as CO and AA having a much longer tail, and YV having the shortest. We should also take a look at the medians by carrier.
```{r}
aggregate(ArrDelay ~ UniqueCarrier, hflights_usable, each(median,mean))
# Get all the summary stats by carrier
aggregate(hflights_usable$ArrDelay, by = list(hflights_usable$UniqueCarrier), FUN = summary)
```

Upon review, we will limit the remainder of our analysis to those airlines with significant delays (the top 3 based on median), for both arrivals and departures.

```{r}
hf_sum <- ddply(hflights_usable, .(UniqueCarrier), summarize, mean=round(mean(ArrDelay),2), max=round(max(ArrDelay),2), median=round(median(ArrDelay,2)))
index <- with(hf_sum, order(-median, -mean, -max))
hf_sum <- hf_sum[index,]
hf_sum <- hf_sum[1:3,]
# Output the resultant so we can identify the carriers
hf_sum
```

So we have identified the carriers with the worst arrival delays as 'OO', 'F9', and 'UA'. Now, let's identify those with the worst departure delays.

```{r}
hf_sum <- ddply(hflights_usable, .(UniqueCarrier), summarize, mean=round(mean(DepDelay),2), max=round(max(DepDelay),2), median=round(median(DepDelay,2)))
index <- with(hf_sum, order(-median, -mean, -max))
hf_sum <- hf_sum[index,]
hf_sum <- hf_sum[1:3,]
```
So we have identified the carriers with the worst departure delays as 'WN', 'CO', and 'UA'. 

Let's work with a smaller data frame and do just a bit more analysis, limiting ourselves to what looks like the worst carrier in terms of delays: UA.
```{r}
hf <- subset(hflights_usable, UniqueCarrier == "UA")

# Now let's compare Destination output
hf_sum <- ddply(hf, .(Dest), summarize, 
                flight_count=length(!is.na(DepDelay)),
                mean_a=round(mean(ArrDelay),2),
                mean_d=round(mean(DepDelay),2), 
                max_a=round(max(ArrDelay),2), 
                max_d=round(max(DepDelay),2), 
                median_a=round(median(ArrDelay),2),
                median_d=round(median(DepDelay),2))
index <- with(hf_sum, order(-flight_count, -median_d, -mean_d, -max_d))
hf_sum <- hf_sum[index,]
hf_sum

# This gives us the top 6 Destination cities in terms of delay (based on median)
# We only include the top 4 based on the very low flight counts of SAN and MCI
hf <- subset(hf, Dest == "SFO" | Dest == "ORD" | Dest == "DEN" | Dest == "IAD")
#One final chart that provides some visual of the delays
ggplot(hf, aes( x = DepDelay)) + geom_histogram() + facet_wrap(~Dest)

# This also shows us some of the summary stats
ddply(hf, .(Dest), summarize, mean=round(mean(ArrDelay),2), max=round(max(ArrDelay),2), median=round(median(ArrDelay,2)), sd=round(sd(ArrDelay),2))
```
Based on all of this we can see that probably the worst destination, when leaving from Houston is San Francisco, a bit surprising given the distance.