---
title: "MSDA R Programming Week 2 Homework"
author: "Ken Markus"
date: "November 6, 2015"
output: html_document
---

The assignment is as follows:
*R has built in functions for choose() and factorial(). Your task is to write your own variation of the choose() function, that does not use either of these functions.*

Our solutions follows by writing two separate functions.

First, we will need a function to calculate the factorial of an integer. Note that we did not include basic error-checking, such as ensuring that the integer was zero or higher.
```{r}
#Calculate the factorial of an integer without leveraging pre-built functions
compute.factorial <- function (x){
  if (x==1) {
    return(x)
  } else {
    return(x * compute.factorial (x-1))
  }
}
```

Following that we must now create the choose function, which is a simple one-line code.
```{r}
#Calculate "n choose r"
compute.choose <- function(n,r) {
  return(compute.factorial(n)/{compute.factorial(r) * compute.factorial(n-r)})
}

```

Ok, great! Now that we have that, let's review the output from the custom function:

```{r}
compute.choose(5,3)
```

and now let's compare it to the in-built function:
```{r}
choose(5,3)
```

Nice! The results should be the same. Let's compare the two for a few more pairs:
```{r}
compute.choose(10,5)
choose(10,5)
compute.choose(5,4)
choose(5,4)
```

So we have seen that `compute.choose` yields output the same as `choose`.

Here is the results of an in-line call of `compute.choose(5,3)`: `r compute.choose(5,3)`
Excellent!
