---
title: "Week 3 Assignment"
author: "Ken Markus"
date: "November 14, 2015"
output: html_document
---

Assignment is summarized as follows:
*Pull data from: https://archive.ics.uci.edu/ml/datasets/Mushroom .*

Take the data, and create a data frame with a subset of the columns (and if you like rows).
```{r}
theUrl="https://archive.ics.uci.edu/ml/machine-learning-databases/mushroom/agaricus-lepiota.data"
#Load all of the data into mushroom_data
mushroom_data <- read.table(file = theUrl, header = FALSE, sep = ",")

#Now, let's make the data more containable, for this hw exercise
md <- mushroom_data[c(1:10),1:4]

# Let's take a look at the output:
md
```

Next add meaningful column names...
```{r}
names(md) <- c("Edible", "cap_shape", "cap_surface", "cap_color")
md
```

Last, replace the abbreviations used in the data
```{r}
require(stringr)
md$Edible <- str_replace(md$Edible,"p", "poisonous")
md$Edible <- str_replace(md$Edible, "e", "edible")

md$cap_shape <- str_replace(md$cap_shape,"b","bell")
md$cap_shape <- str_replace(md$cap_shape, "x", "convex")
md$cap_shape <- str_replace(md$cap_shape, "f", "flat")

md$cap_surface <- str_replace(md$cap_surface, "y$", "scaly")
md$cap_surface <- str_replace(md$cap_surface, "s$", "smooth")

#Note that for the following we needed an exact match
md$cap_color <- str_replace(md$cap_color, "^n$", "brown")
md$cap_color <- str_replace(md$cap_color, "^b$", "buff")
md$cap_color <- str_replace(md$cap_color, "^c$", "cinnamon")
md$cap_color <- str_replace(md$cap_color, "^g$", "gray")
md$cap_color <- str_replace(md$cap_color, "^r$", "green")
md$cap_color <- str_replace(md$cap_color, "^p$", "pink")
md$cap_color <- str_replace(md$cap_color, "^u$", "purple")
md$cap_color <- str_replace(md$cap_color, "^e$", "red")
md$cap_color <- str_replace(md$cap_color, "^w$", "white")
md$cap_color <- str_replace(md$cap_color, "^y$", "yellow")

#Let's tkae a look at the output
md
```

Voila!

While this accomplished the goals of the assignment, I suspect there is a more elegant way to replace the single characters with words, but I was unable to do so unfortunately.